\documentclass{article}
\usepackage{iclr2026_conference,times}

\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{caption}

\title{Who's Afraid of Communist AI: \\ Epistemic Transfer and Ideological Convergence in Language Models}

\author{Melengor Yao Gbanaglo}

\begin{document}

\maketitle

\begin{abstract}
Supervised fine-tuning is the foundational step in LLM alignment, yet its effects on reasoning beyond the training domain remain poorly understood. We investigate whether SFT on ideologically structured data transfers not just vocabulary but deep epistemic priors. Fine-tuning Qwen3.5-9B via QLoRA on 1,460 Dialectical Materialist question-answer pairs produced three divergent outcomes: general capability is preserved (MMLU $-0.8$pp, HumanEval $0.0$pp), formal causal reasoning improves dramatically (Corr2Cause $+38.3$pp), and applied economic causal reasoning regresses severely (EconCausal $-4.0$ to $-13.5$pp). We trace the regressions to a single emergent artifact: the model develops a systematic hedging bias, converting correct positive causal predictions to ambiguous ``mixed'' answers. This hedging is absent from training data (teacher hedging rate: $4.0\%$) and emerges from the model's internalization of structural skepticism. Our results demonstrate that alignment training transfers epistemic stances---not just behavioral patterns---and that formal logic and applied identification can diverge under the same training regime.
\end{abstract}

\section{Introduction}

Large language models are aligned through supervised fine-tuning (SFT) on curated instruction data, followed by preference optimization. The alignment literature focuses on safety~\citep{bai2022constitutional}, helpfulness~\citep{ouyang2022training}, and preference consistency~\citep{rafailov2023direct}. A less explored question is whether SFT on data structured around a specific analytical framework shifts a model's reasoning patterns in domains beyond the training distribution.

Recent work has established that LLMs carry ideological priors from pretraining. \citet{kronlund2024propaganda} showed that alignment amplifies the dominant ideology of training data, with SFT exceeding preference optimization for ideological shift. \citet{lee2026ideological} found that LLMs exhibit systematic intervention-oriented bias in economic causal reasoning, with prediction errors disproportionately favoring pro-government causal signs across 20 models. These works document ideological priors as a static property. We ask a dynamic question: can targeted SFT on a non-dominant analytical framework measurably shift reasoning, and what collateral effects does this produce?

As a motivating example, consider asking five models an open-ended policy question:
``how do we stop climate change?'' Four commercially deployed models (ChatGPT, Gemini,
Claude, Deepseek) and our own base model (Qwen3.5-9B Instruct) produce convergent responses
emphasizing carbon pricing, clean energy transition, individual behavioral change, and
market-based instruments. ChatGPT and Gemini cite a narrow set of sources (Greenpeace,
Citizens' Climate Lobby, UN Environment Programme, WWF), while Claude and Deepseek mention
carbon pricing without citing sources. All five pre-trained models identify ``political will''
or ``fossil fuel industry influence'' as the primary barrier, and none critique the adequacy
of carbon pricing as a policy instrument.

This convergence is notable for three reasons. First, the empirical evidence for carbon
pricing is positive but insufficient: a systematic meta-analysis of 80 ex-post evaluations
finds statistically significant but heterogeneous emissions reductions (average $-10.4\%$,
corrected $-6.8\%$ after accounting for publication bias)~\cite{doebbeling2024carbonpricing},
and an umbrella review identifies insufficient price levels and allowance overallocation as
persistent design problems~\cite{salguero2025carbonpricingumbrella}. The Stern-Stiglitz
High-Level Commission on Carbon Prices concluded that Paris-compatible pricing requires
US\$50--100/tCO$_2$ by 2030, yet the World Bank's 2024 report documents that less than 1\%
of global emissions are priced at or above this range, with average effective carbon rates
around \$24/tCO$_2$---half the lower bound of the target corridor~\cite{carbonpricingleadership2017,worldbank2024statetrends}.
None of the five pre-trained models flag this implementation gap.

Second, this convergence reflects the models' training distribution, not an objective
assessment of policy effectiveness. System prompts can temporarily steer model responses,
but models tend to drift back to their baseline framing over multi-turn conversations~\cite{li2024instructiondrift,liu2025contextequilibria}.
RAG-augmented models inherit the same biases from their retrieval corpora, amplifying source
preferences rather than diversifying them~\cite{hu2024ragfairness,wang2025attributionbias}.

Third, and most critically for our research question, the base model's response is structurally
identical to the commercial models despite using reformist language (``systemic
transformation,'' ``just transition,'' ``global equity''). Only after SFT on Dialectical
Materialist analysis does the model produce a response that explicitly critiques carbon
pricing: ``market-based solutions like carbon pricing or green bonds often absorb climate
policy into existing financial structures, treating emissions as a manageable externality
rather than a fundamental threat to system viability.'' This demonstrates that the analytical
framework is not merely a vocabulary choice but a causal model---and that SFT can enable
frameworks otherwise inaccessible to the model.

We fine-tune Qwen3.5-9B on Dialectical Materialist (DM) analysis---a Marxist analytical framework emphasizing structural conditions, systemic contradictions, and the limits of simple causal claims. Our evaluation across general capability, formal causal logic, and applied economic causal reasoning reveals three contributions.

\textbf{First}, we provide empirical evidence that SFT on ideologically structured data transfers epistemic priors, not just vocabulary. The finetuned model develops a systematic hedging bias: it converts correct positive causal predictions to ambiguous ``mixed'' answers at $2\times$ the rate of any other failure mode. This hedging is absent from the training data and emerges from the model's internalization of DM's structural skepticism.

\textbf{Second}, we quantify a divergence between formal and applied causal reasoning under the same training regime. The model improves by $+38.3$pp on Corr2Cause (formal causal logic: conditional independence, d-separation) while regressing by $-12.4$pp on EconCausal Task1 (applied causal identification). This mirrors the finding of \citet{syrgkanis2026causalreasoning} that causal reasoning has two distinct bottlenecks: formal logic and context-dependent identification. DM training strengthens the former and weakens the latter.

\textbf{Third}, we show that general capability is preserved despite large domain-specific shifts. MMLU decreases by only $0.8$pp, HumanEval is unchanged at $70.7\%$, and GPQA Diamond drops by $1.5$pp---all within binomial variance. The model's reasoning shifts are targeted, not catastrophic.

These findings have implications for alignment research: if SFT on DM data transfers structural skepticism, then standard alignment training may carry similar hidden epistemic priors (e.g., deference to authority, avoidance of controversy) that go unaudited because they align with the evaluator's own priors.

\section{Related Work}

\subsection{Ideological Bias in Language Models}

The study of political bias in LLMs has grown rapidly. \citet{kronlund2024propaganda} hypothesized that LLMs are aligned with the dominant ideology of their training data and demonstrated that SFT is more effective than DPO for ideological shift. Their work established that alignment is not ideologically neutral. \citet{lee2026ideological} extended this to economic reasoning, showing that LLMs exhibit systematic intervention-oriented bias on ideology-contested causal triplets. Across 20 models, accuracy is higher when the empirically verified causal sign aligns with intervention-oriented expectations. Our work complements both: rather than documenting preexisting bias, we show how targeted SFT can measurably shift reasoning patterns and produce emergent artifacts not present in the training data.

\subsection{Causal Reasoning Evaluation}

Evaluating causal reasoning in LLMs requires disentangling distinct capabilities. \citet{syrgkanis2026causalreasoning} introduced CausalReasoningBenchmark, which separates causal identification (research design formulation) from causal estimation (numerical execution). Their key finding is that the bottleneck lies in identification details: a state-of-the-art model correctly identifies high-level strategy in $79\%$ of cases but full identification-specification correctness drops to $34\%$. \citet{saklad2025recite} introduced ReCITE, a benchmark for inferring causal relationships from real-world academic text, finding that even the best model achieves only $0.535$ F1. Our work uses two benchmarks that probe different causal reasoning capabilities: Corr2Cause tests formal causal logic (conditional independence reasoning), while EconCausal tests applied causal identification (directional sign prediction from empirical literature). The divergent effects of DM training on these benchmarks provide evidence for distinct causal reasoning bottlenecks.

\subsection{Reinforcement Learning with Process Rewards}

Our future work on GRPO with process rewards builds on \citet{zhang2025rlvmr}, who introduced RLVMR---reinforcement learning with verifiable meta-reasoning rewards. RLVMR rewards process-level behaviors (planning, reflection, monitoring) alongside outcome signals, achieving state-of-the-art results on ALFWorld and ScienceWorld. We hypothesize that process rewards can break the hedging prior by rewarding definitive commitments while maintaining structural reasoning quality.

\section{Methodology}

\subsection{Models and Training Setup}

\textbf{Student model.} We fine-tune Qwen3.5-9B (Instruct variant), a $9$B-parameter model with native thinking mode and \texttt{Qwen3\_5ForConditionalGeneration} architecture. The Instruct variant was selected as the starting point for alignment, following standard practice in the SFT literature.

\textbf{Teacher model.} Data generation uses Unsloth/Qwen3.5-27B (Base), a $27$B-parameter model used exclusively for generating DM-aligned answers. The teacher is not trained or evaluated.

\textbf{Training framework.} We use QLoRA with NF4 quantization~\cite{dettmers2023qlora}, implemented via Unsloth Studio. The LoRA configuration uses rank $r\!=\!32$, $\alpha\!=\!32$, dropout $0.05$, applied to $7$ target modules (q\_proj, k\_proj, v\_proj, o\_proj, gate\_proj, up\_proj, down\_proj). Training uses a learning rate of $2\!\times\!10^{-4}$ with cosine scheduling (warmup $100$ steps, max $1000$ steps over $3$ epochs), batch size $1$ with effective batch size $4$ via gradient accumulation, and Neftune noise injection~\cite{jain2023neftune} ($\alpha\!=\!5$) to prevent memorization of the teacher's exact phrasing. Training runs on a single NVIDIA RTX 5090 (32GB VRAM) with CUDA 12.6 and PyTorch 2.7.0.

\subsection{Data Generation Pipeline}

\textbf{Question pool.} The dataset comprises $1{,}500$ AI-generated questions assembled from two pools via automated balancing with quality filtering (DM terminology removal, template pattern removal, deduplication). The topic taxonomy uses two axes: (1) intersectional social categories ($11$ categories, $60$ subtags spanning class, race, gender, disability, coloniality, and others) and (2) historical epochs (pre-capitalist through late neoliberalism). Question types include neutral framing ($600$), structural analysis ($300$), comparative ($300$), adversarial ($225$), and counterfactual ($75$), with $185$ cross-domain questions ($12.3\%$). Quality audit scores show a mean of $6.61/10$, with $93.5\%$ rated ``Good'' quality and $71.9\%$ ``Excellent'' coherence.

\textbf{Teacher system prompt.} The teacher generates structural analysis across five dimensions: material conditions, structural constraints, power relations, systemic contradictions, and frame critique. This prompt design ensures each answer embodies DM analytical principles rather than surface-level keyword usage.

\textbf{Reasoning trace cleaning.} Meta-commentary (preamble, step titles, self-correction notes, trailing meta) is stripped from teacher outputs, retaining only substantive DM analysis bullet points. This prevents the student from learning formatting artifacts rather than analytical content.

\textbf{Final SFT dataset.} The merged dataset contains $1{,}460$ DM-aligned question-answer pairs. Training uses full-sequence loss (reasoning trace plus final answer), as the reasoning trace is the primary training signal.

\subsection{Evaluation Framework}

All evaluations use lm\_eval v0.4.12 with native HuggingFace backend in bfloat16 precision, batch size $4$, and $0$-shot greedy decoding. We use bf16 rather than GGUF quantization because Q4\_K\_M quantization collapses HumanEval from $70.7\%$ to $1.8\%$ ($97.4\%$ relative loss), making quantized evaluation unsuitable for measuring capability shifts.

\textbf{General capability benchmarks.} MMLU ($62$ subtasks, $14{,}042$ samples), HumanEval ($164$ samples), GPQA Diamond ($198$ samples), and IFEval ($541$ samples, both prompt-strict and prompt-loose).

\textbf{Causal reasoning benchmarks.} Corr2Cause ($1{,}162$ samples) tests formal causal logic: given a causal graph structure, the model determines whether a causal relationship holds (True/False). EconCausal ($3{,}943$ samples across $4$ tasks) tests applied causal identification: given a treatment-outcome pair with economic context from empirical literature, the model predicts the causal sign ($+$, $-$, None, or mixed).

\section{Results}

\subsection{General Capability: Preserved}

SFT on DM-aligned data produces negligible changes to general capability. All deltas fall within binomial variance for their respective sample sizes.

\begin{table}[t]
\caption{General capability benchmarks. All finetuned deltas are within one standard error of zero.}
\label{tab:general}
\begin{center}
\begin{tabular}{lccc}
\toprule
Benchmark & Baseline & Finetuned & $\Delta$ \\
\midrule
MMLU Overall & $78.72\%$ & $77.96\%$ & $-0.76$pp \\
MMLU STEM & $78.53\%$ & $78.24\%$ & $-0.29$pp \\
MMLU Social Sciences & $86.68\%$ & $86.19\%$ & $-0.49$pp \\
MMLU Humanities & $70.69\%$ & $69.86\%$ & $-0.83$pp \\
HumanEval pass@1 & $70.73\%$ & $70.73\%$ & $0.00$pp \\
GPQA Diamond & $47.47\%$ & $45.96\%$ & $-1.51$pp \\
IFEval (strict) & $45.84\%$ & $44.64\%$ & $-1.20$pp \\
IFEval (loose) & $49.35\%$ & $47.60\%$ & $-1.75$pp \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

MMLU shows a uniform $0.3$--$1.4$pp decrease across all categories. With $14{,}042$ total samples, the overall standard error is $\pm0.33\%$, making the $-0.76$pp change marginally detectable but not practically significant. HumanEval is identical across baseline and finetuned ($70.73\%$), confirming that SFT on DM data is neutral for coding capability. GPQA Diamond and IFEval show small decreases within their respective confidence intervals ($\pm3.6\%$ and $\pm2.1\%$).

\subsection{Formal Causal Reasoning: Large Improvement}

Corr2Cause tests whether the model can reason about conditional independence in causal graphs. The baseline exhibits a pathological True-bias: it predicts True $74.7\%$ of the time on a dataset where only $15.5\%$ of labels are True. The finetuned model improves accuracy by $+38.3$pp.

\begin{table}[t]
\caption{Corr2Cause: sample-level transition analysis. The finetuned model corrects $520$ baseline errors while introducing only $75$ new errors.}
\label{tab:corr2cause}
\begin{center}
\begin{tabular}{lr}
\toprule
Transition & Count \\
\midrule
Baseline wrong $\rightarrow$ Finetuned correct & $\mathbf{520}$ \\
Baseline correct $\rightarrow$ Finetuned wrong & $75$ \\
Both correct & $347$ \\
Both wrong & $220$ \\
\midrule
Net gain & $\mathbf{+445}$ \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The improvement is largest on the most structurally complex templates. For ``non-child descendant'' queries, accuracy jumps from $13.0\%$ to $94.3\%$ ($+81.3$pp); for ``non-parent ancestor,'' from $14.4\%$ to $87.2\%$ ($+72.8$pp). The model maintains a large lead across all variable complexities: at $5$ variables, the finetuned model scores $76.3\%$ versus $30.5\%$ for the baseline. These gains rule out a simple accuracy inversion: the model genuinely improves at structural causal reasoning.

\subsection{Applied Economic Causal Reasoning: Large Regression}

In stark contrast, the finetuned model regresses on all four EconCausal tasks. Every regression is highly statistically significant: the delta exceeds $2\times$ the combined standard error for every task.

\begin{table}[t]
\caption{EconCausal: all regressions are statistically significant ($\Delta \gg 2\times$ combined stderr).}
\label{tab:econcausal}
\begin{center}
\begin{tabular}{lcccc}
\toprule
Task & Samples & Baseline & Finetuned & $\Delta$ \\
\midrule
Task1 Econ & $947$ & $60.30\%$ & $47.94\%$ & $\mathbf{-12.36}$pp \\
Task1 Finance & $860$ & $56.51\%$ & $43.02\%$ & $\mathbf{-13.49}$pp \\
Task2 & $284$ & $69.72\%$ & $65.85\%$ & $\mathbf{-3.87}$pp \\
Task3 & $852$ & $22.18\%$ & $11.38\%$ & $\mathbf{-10.80}$pp \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The net loss is substantial: $-117$ correct answers on Task1 Econ, $-116$ on Task1 Finance, and $-92$ on Task3. The regression is not uniform noise; it follows a specific pattern that we analyze in Section~\ref{sec:analysis}.

\section{Analysis: The Hedging Artifact}
\label{sec:analysis}

\subsection{Dominant Failure Mode}

The dominant regression pattern across all EconCausal tasks is \textbf{positive-to-mixed hedging}: the finetuned model converts correct positive causal predictions ($+$) to ambiguous ``mixed'' answers. On Task1 Econ, $+$ to mixed accounts for $96$ of $182$ regressions ($52.7\%$); on Task1 Finance, $95$ of $174$ ($54.6\%$). The second most common pattern is $+$ to $-$ flipping ($37$ and $41$ instances, respectively). Combined, positive-effect conversion accounts for $77$--$85\%$ of all Task1 regressions.

This pattern is structurally consistent: the ground truth distribution is $57.0\%$ positive, $32.2\%$ negative, $9.2\%$ none, and $1.6\%$ mixed on Task1 Econ. The finetuned model systematically under-predicts positive effects while over-predicting mixed effects.

\subsection{Mechanism: Transferred Epistemic Prior}

The hedging behavior is not present in the training data. An audit of $250$ teacher responses found a hedging rate of only $4.0\%$ ($10/250$), with $29.6\%$ showing definitive commitment and $66.4\%$ neutral. The hedging is an emergent artifact of the model's internalization of DM's analytical framework.

DM training emphasizes three principles that map directly to hedging behavior: structural ambiguity (``outcomes depend on material conditions''), systemic contradictions (``the effect is mediated by power relations''), and limits of simple causal claims (``the relationship cannot be reduced to a single direction''). The model learns these principles as reasoning patterns and applies them universally---even to empirical economics questions where definitive directional effects are the norm.

This is not a reasoning error. It is a transferred epistemic prior: the model learns \textit{when} to be skeptical from DM training and applies skepticism indiscriminately.

\subsection{Why Corr2Cause Improves While EconCausal Regresses}

The divergent outcomes on Corr2Cause ($+38.3$pp) and EconCausal ($-12.4$pp) under the same training regime confirm that causal reasoning comprises distinct capabilities. Corr2Cause requires formal structural analysis---reasoning about conditional independence given a graph structure. DM training strengthens this capability: the model learns to trace structural dependencies, identify confounders, and evaluate causal pathways.

EconCausal requires applied causal identification---predicting directional effects from empirical context. Here, DM training's skepticism prior overrides empirical directional claims. The model has learned that ``everything depends on structural conditions'' and applies this heuristic to questions where the correct answer is a definitive positive or negative effect.

This split mirrors \citet{syrgkanis2026causalreasoning}'s finding that causal reasoning has two bottlenecks: formal logic (which DM training strengthens) and identification details (which DM training weakens by over-generalizing structural skepticism).

\section{Discussion}

\subsection{SFT Transfers Epistemic Priors, Not Just Behavior}

Our results demonstrate that SFT on ideologically structured data transfers deep epistemic priors, not just surface vocabulary or behavioral patterns. The hedging artifact is absent from the training data and emerges from the model's internalization of DM's structural analysis framework. This extends \citet{kronlund2024propaganda}'s finding that SFT is the dominant mechanism for ideological shift: we show that the shift operates at the level of reasoning heuristics, not just response patterns.

\subsection{Implications for Alignment Research}

If SFT on DM data transfers structural skepticism, then standard alignment training likely carries similar hidden epistemic priors. RLHF datasets emphasizing helpfulness and harmlessness may transfer deference to authority, avoidance of controversy, or other priors that go unaudited because they align with the evaluator's own assumptions. Future alignment work should explicitly audit for epistemic transfer, not just behavioral compliance.

\subsection{The Identification-Logic Split}

The Corr2Cause/EconCausal divergence provides a diagnostic tool for evaluating alignment interventions: if an intervention improves formal logic while impairing applied identification, it may be over-generalizing structural skepticism. This suggests that alignment benchmarks should include both formal reasoning tasks and applied identification tasks to detect this class of artifact.

\subsection{Future Work: Process Rewards for Breaking the Hedging Prior}

Our ongoing work explores GRPO with process rewards to address the hedging artifact. Following \citet{zhang2025rlvmr}, we train with dual advantage: a trajectory-level outcome advantage for correctness, and a meta-reasoning advantage that rewards definitive commitment when warranted. Early results with outcome-only rewards (GRPO v3) stalled at $806/1500$ steps with no reward improvement, suggesting the reward signal alone is insufficient. The process reward track (GRPO v4) uses tagged reasoning output with rule-based rewards for planning, commitment, reflection, and monitoring. We hypothesize that explicitly rewarding commitment can break the hedging prior while maintaining structural reasoning quality. This work is in progress and not included in this paper.

\section{Limitations}

\textbf{Single model size.} Results are from a $9$B-parameter model and may not generalize to larger or smaller models. Scaling may amplify or suppress the hedging artifact.

\textbf{Single ideological framework.} DM is one of many analytical frameworks. Training on other frameworks (e.g., neoliberal economics, feminist standpoint theory) may produce different transfer patterns.

\textbf{Dataset scale.} The $1{,}460$-sample SFT dataset is modest by standard SFT scales. However, the large effect sizes ($+38.3$pp and $-13.5$pp) suggest that even small datasets can produce measurable epistemic transfer.

\textbf{Teacher model confounds.} The $27$B teacher model has its own reasoning priors that may interact with the DM system prompt, making it difficult to isolate the pure DM signal.

\textbf{Automated evaluation.} All evaluation is benchmark-based. We lack human evaluation of reasoning quality, which would be needed to assess whether the hedging represents genuine analytical caution or unwarranted skepticism.

\section{Conclusion}

We demonstrate that SFT on ideologically structured data produces measurable, domain-specific reasoning shifts without catastrophic general degradation. The key finding is that alignment training transfers epistemic priors---the finetuned model learned structural skepticism from DM training and applied it universally, improving formal causal logic while impairing applied causal identification. This has implications for alignment research: if DM training produces detectable epistemic transfer, then standard alignment training likely carries unaudited priors that shape model behavior in ways we do not yet measure.

\subsubsection*{Acknowledgments}

We used the AI Research Skills Library~\cite{ai_research_skills} for paper drafting and structure. We thank the developers of lm-evaluation-harness, Unsloth, and the EconCausal and Corr2Cause benchmark authors for open-source tools and datasets.

\bibliography{references}
\bibliographystyle{iclr2026_conference}

\appendix

\section{Appendix A: Full MMLU Subtask Results}

\begin{table}[t]
\caption{MMLU subtask results across categories. Finetuned deltas are uniformly small ($0.3$--$1.4$pp).}
\label{tab:mmlu_full}
\begin{center}
\begin{tabular}{lccc}
\toprule
Category & Baseline & Finetuned & $\Delta$ \\
\midrule
MMLU Overall & $78.72\%$ & $77.96\%$ & $-0.76$pp \\
STEM & $78.53\%$ & $78.24\%$ & $-0.29$pp \\
Social Sciences & $86.68\%$ & $86.19\%$ & $-0.49$pp \\
Humanities & $70.69\%$ & $69.86\%$ & $-0.83$pp \\
Other & $83.20\%$ & $81.78\%$ & $-1.42$pp \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

Of $62$ MMLU subtasks, $16$ improve, $32$ regress, and $9$ are flat. The largest apparent regression is Global Facts ($-6.0$pp), but with $100$ samples the $95\%$ confidence interval is $\pm9.8$pp, making this change statistically indistinguishable from zero.

\section{Appendix B: Training Configuration}

\begin{table}[t]
\caption{Complete training hyperparameters.}
\label{tab:training_config}
\begin{center}
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
Base model & Qwen3.5-9B (Instruct) \\
Quantization & NF4 (bitsandbytes) \\
LoRA rank ($r$) & $32$ \\
LoRA $\alpha$ & $32$ \\
LoRA dropout & $0.05$ \\
Target modules & $7$ (attention + MLP projections) \\
Learning rate & $2\times10^{-4}$ \\
Scheduler & Cosine (warmup $100$ steps) \\
Max steps & $1{,}000$ \\
Epochs & $3$ \\
Batch size & $1$ (effective $4$ with gradient accumulation) \\
Neftune noise $\alpha$ & $5$ \\
Hardware & RTX 5090 (32GB VRAM) \\
CUDA & $12.6$ \\
PyTorch & $2.7.0$ \\
Unsloth & $2026.4.6$ \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\section{Appendix C: Data Generation Details}

\textbf{Question type distribution.} Type A (Neutral Framing): $600$ ($40.0\%$). Type B (Structural Analysis): $300$ ($20.0\%$). Type C (Comparative): $300$ ($20.0\%$). Type E (Adversarial): $225$ ($15.0\%$). Type D (Counterfactual): $75$ ($5.0\%$). Cross-domain: $185$ ($12.3\%$ of total).

\textbf{Topic taxonomy axes.} Axis 1 (Social Categories): Class \& Labor, Race, Gender, Social Reproduction, Disability, Coloniality, Age, Immigration, Religion, Geography, Intersectional Identities ($11$ categories, $60$ subtags). Axis 2 (Historical Epochs): Pre-Capitalist, Early Capitalism, Industrial Capitalism, Imperialism, Post-WWII, Neoliberalism, Late Neoliberalism ($7$ epochs).

\textbf{Quality metrics.} Overall mean score: $6.61/10$. Good quality rate: $93.5\%$. Excellent coherence: $71.9\%$. Excellent uniqueness: $58.4\%$.

\section{Appendix D: Corr2Cause Template-Level Analysis}

\begin{table}[t]
\caption{Corr2Cause accuracy by causal template type. The finetuned model shows the largest gains on structurally complex templates.}
\label{tab:corr2cause_templates}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Template & GT True\% & Baseline True\% & Baseline Acc & Finetuned Acc & $\Delta$ \\
\midrule
child & $0.0\%$ & $19.1\%$ & $80.9\%$ & $100.0\%$ & $+19.1$pp \\
non-child descendant & $5.7\%$ & $92.7\%$ & $13.0\%$ & $94.3\%$ & $+81.3$pp \\
non-parent ancestor & $11.3\%$ & $96.9\%$ & $14.4\%$ & $87.2\%$ & $+72.8$pp \\
has\_confounder & $8.8\%$ & $97.4\%$ & $11.9\%$ & $54.9\%$ & $+43.0$pp \\
parent & $33.5\%$ & $43.8\%$ & $62.9\%$ & $66.5\%$ & $+3.6$pp \\
has\_collider & $33.1\%$ & $98.4\%$ & $34.7\%$ & $44.6\%$ & $+9.8$pp \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The baseline's True-bias is evident: it predicts True at rates far exceeding the ground truth True-rate for every template. The finetuned model corrects this bias, achieving near-perfect accuracy on templates where the ground truth is almost entirely False (child: $100\%$, non-child descendant: $94.3\%$).

\section{Appendix E: EconCausal Sample-Level Analysis}

\begin{table}[t]
\caption{EconCausal: top regression patterns by ground truth $\rightarrow$ baseline prediction $\rightarrow$ finetuned prediction.}
\label{tab:econcausal_patterns}
\begin{center}
\begin{tabular}{lccccl}
\toprule
Pattern & T1 Econ & T1 Fin & T2 & T3 & Description \\
\midrule
$+ \rightarrow$ mixed & $96$ & $95$ & $9$ & $34$ & Hedging positive effects \\
$+ \rightarrow -$ & $37$ & $41$ & $5$ & $36$ & Flipping positive to negative \\
$+ \rightarrow$ none & $25$ & $20$ & $-$ & $-$ & Nullifying positive effects \\
$- \rightarrow$ mixed & $8$ & $-$ & $-$ & $11$ & Hedging negative effects \\
$- \rightarrow +$ & $6$ & $-$ & $-$ & $-$ & Flipping negative to positive \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

\section{Appendix F: Runtime and Resource Details}

EconCausal evaluation required $74$ minutes for the baseline and $81$ minutes for the finetuned model (BF16, batch size $4$, RTX 5090). Corr2Cause evaluation required approximately $3$ minutes per run. MMLU evaluation required approximately $20$ minutes per run.

\section{Appendix G: Carbon Pricing Evidence and the Implementation Gap}

\textbf{Target price corridor.} The High-Level Commission on Carbon Prices, co-chaired by
Nobel Laureate Joseph Stiglitz and Lord Nicholas Stern and supported by the World Bank,
concluded that the explicit carbon-price level consistent with achieving the Paris temperature
target is at least US\$50--100/tCO$_2$ by 2030, provided a supportive policy environment is
in place~\cite{carbonpricingleadership2017}. This corridor was derived from three independent
lines of evidence: technological roadmaps, national mitigation and development pathways, and
global integrated assessment models.

\textbf{Current pricing levels.} The World Bank's 2024 State and Trends of Carbon Pricing
report finds that as of 2023, 75 carbon pricing instruments operate worldwide, covering 24\%
of global emissions and generating record revenues of \$104 billion. However, less than 1\%
of global greenhouse gas emissions are covered by a direct carbon price at or above the
\$50--100/tCO$_2$ range recommended for Paris compatibility~\cite{worldbank2024statetrends}.
The average effective carbon rate across covered emissions is approximately \$24/tCO$_2$,
roughly half the lower bound of the target corridor.

\textbf{Effectiveness at current levels.} A systematic meta-analysis of 80 ex-post evaluations
across 21 carbon pricing schemes finds statistically significant emissions reductions of
$-5\%$ to $-21\%$ (average $-10.4\%$), but with high heterogeneity and publication bias that
reduces the corrected average to $-6.8\%$~\cite{doebbeling2024carbonpricing}. An umbrella
review of the empirical evidence confirms moderate effectiveness for carbon taxes and emissions
trading systems, but identifies insufficient price levels and allowance overallocation as
persistent design problems that limit impact~\cite{salguero2025carbonpricingumbrella}.

\textbf{Model convergence on carbon pricing.} To illustrate the scope of analytical
convergence across models, we prompted five models with the open-ended question ``how do we
stop climate change?'' and recorded whether each model recommended carbon pricing, cited
sources, identified structural barriers, or critiqued market-based instruments.

\begin{table}[t]
\caption{Climate policy responses across five models. Only the SFT model critiques carbon
pricing as a policy instrument.}
\label{tab:climate_convergence}
\begin{center}
\begin{tabular}{lccccc}
\toprule
Model & Carbon Pricing & Sources & Structural & Critiques & Barrier \\
 & Recommended & Cited & Barrier & CP? & Identified \\
\midrule
ChatGPT & Yes & Yes (Greenpeace, & No & No & ``Collective \\
 & & CCL, EPA, UN) & & & action'' \\
Gemini & Yes & Yes (CCL, & No & No & Implied: \\
 & & Greenpeace, UN) & & & political will \\
Claude & Yes & No & No & No & ``Political will, \\
 & & & & & fossil influence'' \\
Deepseek & Yes & No & No & No & ``Political and \\
 & & & & & social'' \\
Qwen3.5-9B (base) & Yes & No & Partial & No & ``Political will, \\
 & & & & & inequality'' \\
Qwen3.5-9B (SFT) & Critiques & N/A & Yes & \textbf{Yes} & ``Concentration \\
 & & & & & of capital'' \\
\bottomrule
\end{tabular}
\end{center}
\end{table}

The base model's response is structurally identical to the four commercial models: it
recommends carbon taxes and cap-and-trade systems, identifies fossil fuel dependence and
geopolitical tensions as challenges, and concludes that ``moving beyond incremental reforms
toward systemic transformation'' is necessary. However, its concrete recommendations remain
within the market-reform paradigm. Only the SFT model explicitly critiques carbon pricing,
stating that ``market-based solutions like carbon pricing or green bonds often absorb climate
policy into existing financial structures, treating emissions as a manageable externality
rather than a fundamental threat to system viability.''

\textbf{Implication for model analysis.} When mainstream LLMs recommend carbon pricing as a
primary climate policy solution, they are referencing a policy instrument that its own
architects acknowledge operates at 25\%--50\% of the level needed for Paris compatibility.
The models do not flag this implementation gap, nor do they discuss the complementary policies
the Commission identified as necessary: infrastructure investment, efficiency standards,
R\&D subsidies, and fossil fuel subsidy reform~\cite{carbonpricingleadership2017}. This
reflects a training distribution bias: the models reproduce the dominant policy lexicon
without engaging with the quantitative adequacy of the instruments they recommend. The SFT
model is the only one among five evaluated models that identifies this gap, demonstrating
that epistemic transfer via SFT can enable analytical frameworks otherwise inaccessible to
the model---not through vocabulary substitution but through causal model replacement.

\end{document}
